---
title: "Take-Home_Ex02"
date: "13 Feb 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

# Overview

## Context

As a result of the COVID-19 outbreak in Indonesia, a widespread immunization campaign is currently being conducted there. The program was launched on January 13, 2021, when President Joko Widodo had his vaccination at the presidential palace. Indonesia is ranked third in Asia and fifth overall for total dosages administered.

Wikipedia states that as of February 5, 2023, at 18:00 WIB (UTC+7), 204,266,655 individuals had received the first dose of the vaccine and 175,131,893 individuals had received the full course of vaccination; 69,597,474 of these individuals had received the booster dose or third dose, and 1,585,164 had received the fourth dose. With 103.46% of the population fully immunized, Jakarta has the highest rate, followed by Bali and the Special Region of Yogyakarta with 85.45% and 83.02%, respectively.

## Objectives

Exploratory Spatial Data Analysis (ESDA) hold tremendous potential to address complex problems facing society. In this study, you are tasked to apply appropriate Local Indicators of Spatial Association (LISA) and Emerging Hot Spot Analysis (EHSA) to undercover the spatio-temporal trends of COVID-19 vaccination in DKI Jakarta.

## Task

Choropleth Mapping and Analysis

-   Compute the monthly vaccination rate from July 2021 to June 2022 at sub-district (also known as kelurahan in Bahasa Indonesia) level,

-   Prepare the monthly vaccination rate maps by using appropriate tmap functions,

-   Describe the spatial patterns revealed by the choropleth maps (not more than 200 words).

## Data used

| Type       | Name                                     | Format | Description                   | Source                                   |
|---------------|---------------|---------------|---------------|---------------|
| Aspatial   | Riwayat File Vaksinasi DKI Jakarta       | csv    | Daily vaccination data        | Open Data Vaksinasi Provinsi DKI Jakarta |
| Geospatial | DKI Jakarta administration boundary 2019 | SHP    | geoBoundaries data of Jakarta | Indonesia Geospatial                     |

: Table 1: Data used

For the Aspatial Data, I decided to use the dataset available on 1st Sept 2022 for this exercise. 

## Packages used

For this exercise, we'll be using the following packages:

-   sf: Manage and process vector-based geospatial data in R

-   sfdep: 

-   maptools: Convert Spatial objects into ppp format of spatstat

-   tmap: Plotting cartographic quality static point patterns maps or interactive maps by using leaflet API

-   tidyverse: a collection of functions, data, and documentation that extends the capabilities of base R.

-   knitr: 

-   readxl: The readxl package makes it easy to get data out of Excel and into R

We will be using p_load function of pacman package to install and load required packages.

```{r}
pacman::p_load(sf, sfdep, tmap, maptools, tidyverse, knitr, readxl)
```

# Import Data

## Importing Geospatial

::: panel-tabset

### Import
```{r}
jkt <- st_read(dsn = "data/geospatial/BATAS DESA DESEMBER 2019 DUKCAPIL DKI JAKARTA", 
                layer = "BATAS_DESA_DESEMBER_2019_DUKCAPIL_DKI_JAKARTA")
```

### Glimpse
```{r}
glimpse(jkt)
```

::: 

## Importing Aspatial

::: panel-tabset

### Import
```{r}
vaccinated <- read_excel("data/aspatial/Data Vaksinasi Berbasis Kelurahan (01 September 2022).xlsx")
```

### Glimpse
```{r}
glimpse(vaccinated)
```

:::

## Perform relational join
```{r}
# jkt_2022 <- left_join(jkt,hunan2012)%>%
#   select(1:4, 7, 15)
```


