---
title: "In-class_Ex06"
date: "13 Feb 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

# Overview

## Install and load packages
```{r}
pacman::p_load(sf, sfdep, tmap, tidyverse)
```

# DATA

## Import Geospatial Data
```{r}
hunan <- st_read(dsn = "data/geospatial", 
                 layer = "Hunan")
```


## Import Aspatial Data
```{r}
hunan2012 <- read_csv("data/aspatial/Hunan_2012.csv")
hunan2012
```


## Combine both df using left-join

Always do left-join cause its best to put the sf df the left side then right side is the tibble

```{r}
hunan_GDPPC <- left_join(hunan, hunan2012) |>
  select(1:4, 7, 15)
```

NOTE: during joining,must have a common field (unique identifier) then another thing to take note is to make sure both columns have same number of attributes(?)


## Plotting Chrorpleth map
```{r}
tmap_mode("plot")
tm_shape(hunan_GDPPC)+
  tm_fill("GDPPC", 
          style = "quantile", 
          palette = "Blues",
          title = "GDPPC") +
  tm_layout(main.title = "Distribution of GDP per capita by district",
            main.title.position = "center",
            main.title.size = 1.0,
            legend.height = 0.45, 
            legend.width = 0.35,
            frame = TRUE) +
  tm_borders(alpha = 0.5) +
  tm_compass(type="8star", size = 2) +
  tm_scale_bar() +
  tm_grid(alpha =0.2) 
```


# Identify neighbours

## Contibuity neighbours method
.before = 1 is to put the newly created column at the first column onwards.
```{r}
cn_queen <- hunan_GDPPC |>
  mutate(nb = st_contiguity(geometry), .before = 1)
```



## Neighbour list by Rook's method
queen = false is to know that the default queen is false then its rook.
```{r}
cn_rook <- hunan_GDPPC |>
  mutate(nb = st_contiguity(geometry), queen = FALSE, .before = 1)

```


```{r}
wm_q = hunan_GDPPC |>
  mutate(nb = st_contiguity(geometry), 
         wt = st_weights(nb), 
         .before = 1)
```



## Contiguity rook method
```{r}
wm_r = hunan_GDPPC |>
  mutate(nb = st_contiguity(geometry), 
         queen = FALSE,
         wt = st_weights(nb), 
         .before = 1)
```

