---
title: "In-class_Ex12"
date: "3 April 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  warning: false
editor: visual
---

Link to in class ex11, starting from calibrating the spatial interaction variable

```{r}
pacman::p_load(tmap, sf, sp, stplanr, 
               reshape2, broom, performance,
               gtsummary, ggpubr, tidyverse)
```

## read rds
```{r}
SIM_data <- read_rds("data/SIM_data/SIM_data.rds")
```

## ggplot
```{r}
ggplot(data = SIM_data,
       aes(x = TRIPS)) + 
  geom_histogram()
```


```{r}
ggplot(data = SIM_data,
       aes(x = dist,
           y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r}
ggplot(data = SIM_data,
       aes(x = log(dist),
           y = log(TRIPS))) +
  geom_point() +
  geom_smooth(method = lm)
```

```{r}
summary(SIM_data)
```

```{r}
SIM_data$DESTIN_AGE7_12 <- ifelse(
  SIM_data$DESTIN_AGE7_12 == 0,
  0.99, SIM_data$DESTIN_AGE7_12)
SIM_data$DESTIN_AGE13_24 <- ifelse(
  SIM_data$DESTIN_AGE13_24 == 0,
  0.99, SIM_data$DESTIN_AGE13_24)
SIM_data$DESTIN_AGE25_64 <- ifelse(
  SIM_data$DESTIN_AGE25_64 == 0,
  0.99, SIM_data$DESTIN_AGE25_64)
SIM_data$ORIGIN_AGE7_12 <- ifelse(
  SIM_data$ORIGIN_AGE7_12 == 0,
  0.99, SIM_data$ORIGIN_AGE7_12)
SIM_data$ORIGIN_AGE13_24 <- ifelse(
  SIM_data$ORIGIN_AGE13_24 == 0,
  0.99, SIM_data$ORIGIN_AGE13_24)
SIM_data$ORIGIN_AGE25_64 <- ifelse(
  SIM_data$ORIGIN_AGE25_64 == 0,
  0.99, SIM_data$ORIGIN_AGE25_64)
```


```{r}
summary(SIM_data)
```
# Unconstrained Spatial Interaction Model

```{r}
uncSIM <- glm(formula = TRIPS ~ 
                log(ORIGIN_AGE25_64) + 
                log(DESTIN_AGE25_64) +
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
uncSIM
```

## R-squared function

Its not from the possion classification, mannually calculated, not real calculations of R squared.

```{r}
CalcRSquared <- function(observed,estimated){
  r <- cor(observed,estimated)
  R2 <- r^2
  R2
}
```

```{r}
CalcRSquared(uncSIM$data$TRIPS, uncSIM$fitted.values)
```


```{r}
r2_mcfadden(uncSIM)
```

## Origin (Production) constrained SIM

```{r}
orcSIM <- glm(formula = TRIPS ~ 
                 ORIGIN_SZ +
                 log(DESTIN_AGE25_64) +
                 log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(orcSIM)
```


```{r}
CalcRSquared(orcSIM$data$TRIPS, orcSIM$fitted.values)
```
## Destination constrained
```{r}
decSIM <- glm(formula = TRIPS ~ 
                DESTIN_SZ + 
                log(ORIGIN_AGE25_64) + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(decSIM)
```

```{r}
CalcRSquared(decSIM$data$TRIPS, decSIM$fitted.values)
```

## Doubly constrained
```{r}
dbcSIM <- glm(formula = TRIPS ~ 
                ORIGIN_SZ + 
                DESTIN_SZ + 
                log(dist),
              family = poisson(link = "log"),
              data = SIM_data,
              na.action = na.exclude)
summary(dbcSIM)
```

```{r}
CalcRSquared(dbcSIM$data$TRIPS, dbcSIM$fitted.values)
```


## Model comparison
```{r}
model_list <- list(unconstrained=uncSIM,
                   originConstrained=orcSIM,
                   destinationConstrained=decSIM,
                   doublyConstrained=dbcSIM)
```

```{r}
compare_performance(model_list,
                    metrics = "RMSE")
```


## Visualising fitted
```{r}
df <- as.data.frame(uncSIM$fitted.values) %>%
  round(digits = 0)
```

```{r}
SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(uncTRIPS = "uncSIM$fitted.values")
```

```{r}
df <- as.data.frame(orcSIM$fitted.values) %>%
  round(digits = 0)
```

```{r}
SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(orcTRIPS = "orcSIM$fitted.values")
```


```{r}
df <- as.data.frame(decSIM$fitted.values) %>%
  round(digits = 0)
```


```{r}
SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(decTRIPS = "decSIM$fitted.values")
```


```{r}
df <- as.data.frame(dbcSIM$fitted.values) %>%
  round(digits = 0)
```

```{r}
SIM_data <- SIM_data %>%
  cbind(df) %>%
  rename(dbcTRIPS = "dbcSIM$fitted.values")
```



```{r}
unc_p <- ggplot(data = SIM_data,
                aes(x = uncTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

orc_p <- ggplot(data = SIM_data,
                aes(x = orcTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

dec_p <- ggplot(data = SIM_data,
                aes(x = decTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

dbc_p <- ggplot(data = SIM_data,
                aes(x = dbcTRIPS,
                    y = TRIPS)) +
  geom_point() +
  geom_smooth(method = lm)

ggarrange(unc_p, orc_p, dec_p, dbc_p,
          ncol = 2,
          nrow = 2)
```


Confirmed the double constrainted model is the better model by giving a smaller root mean squared error. 











