---
title: "Hands-on_Ex03"
date: "27 Jan 2023"
date-modified: "`r Sys.Date()`"
number-sections: true
format: html
execute: 
  echo: true
  eval: true
  message: true
  warning: false
editor: visual
---

# Install and load packages

```{r}
pacman::p_load(maptools, sf, raster, spatstat, tmap)
```

## Spatial Data Wrangling - Importing the spatial data

```{r}
childcare_sf <- st_read("data/child-care-services-geojson.geojson") %>%
  st_transform(crs = 3414)
```

```{r}
sg_sf <- st_read(dsn = "data", layer="CostalOutline")
```

```{r}
mpsz_sf <- st_read(dsn = "data", 
                layer = "MP14_SUBZONE_WEB_PL")
```

## Mapping the geospatial data sets

```{r}
tmap_mode('view')
tm_shape(childcare_sf) + tm_dots()
```

```{r}
tmap_mode('plot')
```

### Geospatial Data wrangling #Converting sf data frames to sp's Spatial\* class

```{r}
childcare <- as_Spatial(childcare_sf)
mpsz <- as_Spatial(mpsz_sf)
sg <- as_Spatial(sg_sf)
```

```{r}
childcare
```

```{r}
mpsz
```

```{r}
sg
```

### Converting the Spatial\* class into generic sp format

```{r}
childcare_sp <- as(childcare, "SpatialPoints")
sg_sp <- as(sg, "SpatialPolygons")
```

```{r}
childcare_sp
```

```{r}
sg_sp
```

##Converting the generic sp format into spatstat's ppp format

```{r}
childcare_ppp <- as(childcare_sp, "ppp")
childcare_ppp
```

```{r}
plot(childcare_ppp)
```

```{r}
summary(childcare_ppp)
```

### Handling duplicated points

```{r}
any(duplicated(childcare_ppp))
```

```{r}
multiplicity(childcare_ppp)
```

```{r}
sum(multiplicity(childcare_ppp) > 1)
```

```{r}
tmap_mode('view')
tm_shape(childcare) +
  tm_dots(alpha=0.4, 
          size=0.05)
```

```{r}
tmap_mode('plot')
```

#### jittering approach

```{r}
childcare_ppp_jit <- rjitter(childcare_ppp, 
                             retry=TRUE, 
                             nsim=1, 
                             drop=TRUE)
```

```{r}
any(duplicated(childcare_ppp_jit))
```

### Creating owin object

```{r}
sg_owin <- as(sg_sp, "owin")
```

```{r}
plot(sg_owin)
```

```{r}
summary(sg_owin)
```

### Combining point events object and owin object

```{r}
childcareSG_ppp = childcare_ppp[sg_owin]
```

```{r}
summary(childcareSG_ppp)
```

# First-order Spatial Point Patterns Analysis

#### #Kernel Density Estimation \# Computing kernel density estimation using automatic bandwidth selection method

```{r}
kde_childcareSG_bw <- density(childcareSG_ppp,
                              sigma=bw.diggle,
                              edge=TRUE,
                            kernel="gaussian") 
```

```{r}
plot(kde_childcareSG_bw)
```

```{r}
bw <- bw.diggle(childcareSG_ppp)
bw
```

#### Rescalling KDE values

```{r}
childcareSG_ppp.km <- rescale(childcareSG_ppp, 1000, "km")
```

```{r}
kde_childcareSG.bw <- density(childcareSG_ppp.km, sigma=bw.diggle, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG.bw)
```

### Working with different automatic bandwidth methods

```{r}
 bw.CvL(childcareSG_ppp.km)
```

```{r}
bw.scott(childcareSG_ppp.km)
```

```{r}
bw.ppl(childcareSG_ppp.km)
```

```{r}
bw.diggle(childcareSG_ppp.km)
```

```{r}
kde_childcareSG.ppl <- density(childcareSG_ppp.km, 
                               sigma=bw.ppl, 
                               edge=TRUE,
                               kernel="gaussian")
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "bw.diggle")
plot(kde_childcareSG.ppl, main = "bw.ppl")
```

### Working with different kernel methods

```{r}
par(mfrow=c(2,2))
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="gaussian"), 
     main="Gaussian")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="epanechnikov"), 
     main="Epanechnikov")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="quartic"), 
     main="Quartic")
plot(density(childcareSG_ppp.km, 
             sigma=bw.ppl, 
             edge=TRUE, 
             kernel="disc"), 
     main="Disc")
```

### Computing KDE by using fixed bandwidth

```{r}
kde_childcareSG_600 <- density(childcareSG_ppp.km, sigma=0.6, edge=TRUE, kernel="gaussian")
plot(kde_childcareSG_600)
```

### Computing KDE by using adaptive bandwidth

```{r}
kde_childcareSG_adaptive <- adaptive.density(childcareSG_ppp.km, method="kernel")
plot(kde_childcareSG_adaptive)
```

```{r}
par(mfrow=c(1,2))
plot(kde_childcareSG.bw, main = "Fixed bandwidth")
plot(kde_childcareSG_adaptive, main = "Adaptive bandwidth")
```

### Converting KDE output into grid object

```{r}
gridded_kde_childcareSG_bw <- as.SpatialGridDataFrame.im(kde_childcareSG.bw)
spplot(gridded_kde_childcareSG_bw)
```

### Converting gridded output into raster

```{r}
kde_childcareSG_bw_raster <- raster(gridded_kde_childcareSG_bw)
```

```{r}
kde_childcareSG_bw_raster
```

### Assigning projection systems

```{r}
projection(kde_childcareSG_bw_raster) <- CRS("+init=EPSG:3414")
kde_childcareSG_bw_raster
```

### Visualising the output in tmap

```{r}
tm_shape(kde_childcareSG_bw_raster) + 
  tm_raster("v") +
  tm_layout(legend.position = c("right", "bottom"), frame = FALSE)
```

### Comparing Spatial Point Patterns using KDE #Extracting study area

```{r}
pg = mpsz[mpsz@data$PLN_AREA_N == "PUNGGOL",]
tm = mpsz[mpsz@data$PLN_AREA_N == "TAMPINES",]
ck = mpsz[mpsz@data$PLN_AREA_N == "CHOA CHU KANG",]
jw = mpsz[mpsz@data$PLN_AREA_N == "JURONG WEST",]
```

```{r}
par(mfrow=c(2,2))
plot(pg, main = "Ponggol")
plot(tm, main = "Tampines")
plot(ck, main = "Choa Chu Kang")
plot(jw, main = "Jurong West")
```

### Converting the spatial point data frame into generic sp format

```{r}
pg_sp = as(pg, "SpatialPolygons")
tm_sp = as(tm, "SpatialPolygons")
ck_sp = as(ck, "SpatialPolygons")
jw_sp = as(jw, "SpatialPolygons")
```

### Creating owin object

```{r}
pg_owin = as(pg_sp, "owin")
tm_owin = as(tm_sp, "owin")
ck_owin = as(ck_sp, "owin")
jw_owin = as(jw_sp, "owin")
```

### Combining childcare points and the study area

```{r}
childcare_pg_ppp = childcare_ppp_jit[pg_owin]
childcare_tm_ppp = childcare_ppp_jit[tm_owin]
childcare_ck_ppp = childcare_ppp_jit[ck_owin]
childcare_jw_ppp = childcare_ppp_jit[jw_owin]
```

```{r}
childcare_pg_ppp.km = rescale(childcare_pg_ppp, 1000, "km")
childcare_tm_ppp.km = rescale(childcare_tm_ppp, 1000, "km")
childcare_ck_ppp.km = rescale(childcare_ck_ppp, 1000, "km")
childcare_jw_ppp.km = rescale(childcare_jw_ppp, 1000, "km")
```

```{r}
par(mfrow=c(2,2))
plot(childcare_pg_ppp.km, main="Punggol")
plot(childcare_tm_ppp.km, main="Tampines")
plot(childcare_ck_ppp.km, main="Choa Chu Kang")
plot(childcare_jw_ppp.km, main="Jurong West")
```

### Computing KDE

```{r}
par(mfrow=c(2,2))
plot(density(childcare_pg_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tempines")
plot(density(childcare_ck_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Choa Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=bw.diggle, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
```

### Computing fixed bandwidth KDE

```{r}
par(mfrow=c(2,2))
plot(density(childcare_ck_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Chou Chu Kang")
plot(density(childcare_jw_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="JUrong West")
plot(density(childcare_pg_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Punggol")
plot(density(childcare_tm_ppp.km, 
             sigma=0.25, 
             edge=TRUE, 
             kernel="gaussian"),
     main="Tampines")
```

## Nearest Neighbour Analysis

### Testing spatial point patterns using Clark and Evans Test

```{r}
clarkevans.test(childcareSG_ppp,
                correction="none",
                clipregion="sg_owin",
                alternative=c("clustered"),
                nsim=99)
```

#### Clark and Evans Test: Choa Chu Kang planning area

```{r}
clarkevans.test(childcare_ck_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

#### Clark and Evans Test: Tampines planning area

```{r}
clarkevans.test(childcare_tm_ppp,
                correction="none",
                clipregion=NULL,
                alternative=c("two.sided"),
                nsim=999)
```

# Second Order Spatial Point Patterns Analysis Methods

## Analysing Spatial Point Process Using G-Function

### Choa Chu Kang planning area Computing G-function estimation

```{r}
G_CK = Gest(childcare_ck_ppp, correction = "border")
plot(G_CK, xlim=c(0,500))
```

### Performing Complete Spatial Randomness Test

```{r}
G_CK.csr <- envelope(childcare_ck_ppp, Gest, nsim = 999)
```

```{r}
plot(G_CK.csr)
```

### Tampines planning area - Computing G-function estimation

```{r}
G_tm = Gest(childcare_tm_ppp, correction = "best")
plot(G_tm)
```

### Performing Complete Spatial Randomness Test

```{r}
G_tm.csr <- envelope(childcare_tm_ppp, Gest, correction = "all", nsim = 999)
```

```{r}
plot(G_tm.csr)
```

## Analysing Spatial Point Process Using F-Function

### Choa Chu Kang planning area

```{r}
F_CK = Fest(childcare_ck_ppp)
plot(F_CK)
```

### Performing Complete Spatial Randomness Test Monte Carlo test with F-fucntion

```{r}
F_CK.csr <- envelope(childcare_ck_ppp, Fest, nsim = 999)
```

```{r}
plot(F_CK.csr)
```

### Tampines planning area Computing F-function estimation

```{r}
F_tm = Fest(childcare_tm_ppp, correction = "best")
plot(F_tm)
```

### Performing Complete Spatial Randomness Test

```{r}
F_tm.csr <- envelope(childcare_tm_ppp, Fest, correction = "all", nsim = 999)
```

```{r}
plot(F_tm.csr)
```

## Analysing Spatial Point Process Using K-Function

### Choa Chu Kang planning area Computing K-fucntion estimate

```{r}
K_ck = Kest(childcare_ck_ppp, correction = "Ripley")
plot(K_ck, . -r ~ r, ylab= "K(d)-r", xlab = "d(m)")
```

### Performing Complete Spatial Randomness Test

```{r}
K_ck.csr <- envelope(childcare_ck_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_ck.csr, . - r ~ r, xlab="d", ylab="K(d)-r")
```

### Tampines planning area Computing K-function estimation

```{r}
K_tm = Kest(childcare_tm_ppp, correction = "Ripley")
plot(K_tm, . -r ~ r, 
     ylab= "K(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

### Performing Complete Spatial Randomness Test

```{r}
K_tm.csr <- envelope(childcare_tm_ppp, Kest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(K_tm.csr, . - r ~ r, 
     xlab="d", ylab="K(d)-r", xlim=c(0,500))
```

## Analysing Spatial Point Process Using L-Function

### Choa Chu Kang planning area Computing L Fucntion estimation

```{r}
L_ck = Lest(childcare_ck_ppp, correction = "Ripley")
plot(L_ck, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)")
```

### Performing Complete Spatial Randomness Test

```{r}
L_ck.csr <- envelope(childcare_ck_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_ck.csr, . - r ~ r, xlab="d", ylab="L(d)-r")
```

### Tampines planning area Computing L-function estimate

```{r}
L_tm = Lest(childcare_tm_ppp, correction = "Ripley")
plot(L_tm, . -r ~ r, 
     ylab= "L(d)-r", xlab = "d(m)", 
     xlim=c(0,1000))
```

### Performing Complete Spatial Randomness Test

```{r}
L_tm.csr <- envelope(childcare_tm_ppp, Lest, nsim = 99, rank = 1, glocal=TRUE)
```

```{r}
plot(L_tm.csr, . - r ~ r, 
     xlab="d", ylab="L(d)-r", xlim=c(0,500))
```
